A boundary-integral framework to simulate \\ viscous erosion of a porous medium

### 2018-04-14 docs/methods/erosion_methods.tex
NM

This BIE formulation requires that $\bd\Omega$ is smooth.  Unfortunately, smoothness is not guaranteed by erosion since corners develop at locations where the shear stress vanishes.  One option is to modify our BIE to handle corners~\cite{rac-ser2017, ser-rok2016, hel2011, gil-hao-mar2014, bre2012}, but these methods require the corner's location, a non-uniform spaced mesh, or are ill-conditioned.  Instead, make a slight modification to the erosion model in Section~\ref{sec:thetaL} which allows us to control the smoothness of the grains.  
\todo[inline]{There is a typo in the above sentence (missing 'we'?). Also, technically there is  a dangling participle here; i.e.~is it the erosion model which is in Section 2.3 or the modification? A slight rephrasing might clarify.}
In this manner, the BIE as described above can be applied.

@@

We do not employ an FMM to evaluate the shear stress at $\xx_i$, $i=1,\ldots,N$, so the cost of evaluating the shear stress is $\bigO(N^2)$.  However, this calculation is only performed once per time step as opposed to the velocity double-layer potential, where we do employ the FMM, that must be evaluated at each GMRES iteration.  In future work, we plan to use a kernel-independent FMM~\cite{yin-bir-zor2004} to evaluate the shear stress with $\bigO(N)$ operations.

@@

First, we must address a second potential source of numerical instability. The erosion law~\eqref{eqn:Vn} depends on the {\em absolute value} of shear stress, $\atau$, which exhibits a corner anywhere $\tau$ changes sign, i.e.~at a stagnation point. The corner in $\atau$ introduces high frequencies into~\eqref{eqn:thetaevo1} which could result in either numerical instability or an extremely stringent step-size restriction. To avoid these difficulties, we smooth $\atau$ via a narrow Gaussian filter of width $\sigma$
\begin{align}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma}
 \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha' \, ,
\end{align}
with corresponding normal and tangential interface velocities
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig} \, .
\end{align}
This smoothing further suppresses the feedback of high frequencies between $\atau$ and the body shape.

@@

These fixed-area simulations enable us to precisely characterize the changing geometry as it approaches the terminal state. In particular, we measure the aspect ratio of the body (the width to height) and the opening angle of the front/rear corners. In these tests, we keep the spatial resolution fixed, $N_\iin = 1024$, and show three different cases of the smoothing parameters, $\eps$ and $\sigma$, to assess how the numerical regularization influences shape. In the three cases, we keep $\eps = \sigma$ always and set them to the values 20/1024, 10/1024, and 5/1024. Loosely, these values correspond to smoothing over the nearest 20, 10, and 5 grid points, respectively. Figure~\ref{fig:arangle}(a) shows the aspect ratio as it varies in time. The initially circular geometry has an aspect ratio of unity, and, as erosion carves a more slender form, the aspect ratio increases monotonically. The three simulations are all very close to one another, demonstrating convergence as $\eps, \sigma \to 0$. In the simulation with these least smoothing ($\eps = \sigma = 5/1024$), the aspect ratio tends to a value of 2.7 in long time.

@@

\subsection{Convergence test}
As a final validation, we now verify that our method converges with the expected second-order accuracy in time. For this, we run the same single-body simulation shown in Figs.~\ref{01bodseq} and~\ref{shrink_intface} (except with a coarser $N_\iin=256$) to a stopping time of $t_s = 10^{-2}$. This stopping time is 56\% of the vanishing time, and was selected so that the body will have eroded substantially but is not small enough to compromise accuracy (due to the stress diverging).  We assess error by measuring the $L^2$-difference in the body shape obtained from simulations of successive resolution, with $\Delta t$ halved at each stage. In Table~\ref{convtab}, we report these errors along with the order of accuracy, obtained by Richardson-extrapolation-type calculations. The table confirms that our method converges with expected second-order accuracy. The last column of the table shows the CPU time taken for the simulation to run on a single-processor machine, demonstrating a nearly linear increase in computational time. Though the main parts of the algorithm are easily paralyzed, the single-processor times provide a simple assessment of method's complexity.

@@

Looking back at the 50-body simulation in Fig.~\ref{fig:50bodies}, many of these same features can be observed. Bodies tends to develop corner-like features connected by flat faces, which causes them to appear polygonal. Nearly straight channels form between the bodies, and a few of these appear to be preferred channels which transmit the most flow and therefore open the fastest.

@@

To further quantify these observations, Fig.~\ref{fig:mbodyplots} shows several measurements taken from the multibody simulations. In all cases, the black curves correspond to the above 6-body simulation and gray curves to the 50-body simulation from Fig.~\ref{fig:50bodies}. First, Fig.~\ref{fig:mbodyplots}(a) shows the fraction of area occupied by the solid bodies as it varies in time. In the 6-body simulation, the initial configuration comprises about 40\% of the domain area, and this percentage steadily decreases as the bodies erode. The 50-body simulation shows a similar trend.

Next, Figure~\ref{fig:mbodyplots}(b) shows how the flow rate, $\umax$, changes in time as a result of fixing the pressure drop across the cell. Notice that, in this figure, the vertical axis is logarithmic. Both simulations show the flow rate to increase rapidly, even during the early stages of erosion when the area fraction has not changed that significantly. Due to the logarithmic axis, the nearly linear trend seen in the figure corresponds to exponential growth of $\umax$. The change in area-fraction alone cannot account for this rapid growth, suggesting the reshaping process to be vital. In particular, the formation of channels observed in Fig.~\ref{fig:06bodies} may account for these higher flow rates.

Naturally, the flow allowed by the medium is connected to the amount of drag the bodies exert. We therefore show in Fig.~\ref{fig:mbodyplots}(c) the total drag exerted by all bodies as it varies in time. In both cases, the drag remains nearly constant throughout much of the simulation and then drops to zero in the final moments. Although all bodies are diminishing in size, the associated increase in $\umax$ leads to the nearly constant drag that is observed here.

>>>>

This BIE formulation requires that $\bd\Omega$ is smooth.  Unfortunately, smoothness is not guaranteed by erosion since corners develop at locations where the shear stress vanishes.  One option is to modify our BIE to handle corners~\cite{rac-ser2017, ser-rok2016, hel2011, gil-hao-mar2014, bre2012}, but these methods require the corner's location, a non-uniform spaced mesh, or are ill-conditioned.  Instead, make a slight modification to the erosion model in Section~\ref{sec:thetaL} which allows us to control the smoothness of the grains.  
\todo[inline]{There is a typo in the above sentence (missing 'we'?). Also, technically there is  a dangling participle; i.e.~is it the erosion model which is in Section 2.3 or the modification? A slight rephrasing might clarify.}
In this manner, the BIE as described above can be applied.

@@

We do not employ an FMM to evaluate the shear stress at $\xx_i$, $i=1,\ldots,N$, so the cost of evaluating the shear stress is $\bigO(N^2)$.  However, this calculation is only performed once per time step as opposed to the velocity double-layer potential, where we do employ the FMM, that must be evaluated at each GMRES iteration. 
\todo[inline]{The above sentence is confusing. Try to rephrase and shorten it.}
In future work, we plan to use a kernel-independent FMM~\cite{yin-bir-zor2004} to evaluate the shear stress with $\bigO(N)$ operations.

@@

First, we must address another potential source of numerical instability. The erosion law~\eqref{eqn:Vn} depends on the {\em absolute value} of shear stress, $\atau$, which exhibits a corner anywhere $\tau$ changes sign, i.e.~at a stagnation point. The corner in $\atau$ introduces high frequencies into~\eqref{eqn:thetaevo1} which could result in either numerical instability or an extremely stringent step-size restriction. To avoid these difficulties, we smooth $\atau$ via a narrow Gaussian filter of width $\sigma \ll 1$
\begin{align}
\atausig = \int_{\gamma} \frac{1}{\sqrt{2\pi} \sigma}
 \exp \left(- \frac{\alpha'^2}{2 \sigma^2}\right) \abs{ \tau(\alpha - \alpha') } \, d\alpha' \, ,
\end{align}
with corresponding normal and tangential interface velocities
\begin{align}
& \Vnsig = \atausig +  \eps \mean{\atausig}
\left(\pderiv{\theta}{\alpha} - 1 \right), \\
& \pderiv{\Vssig}{\alpha} = \thalpha \Vnsig - \mean{\thalpha \Vnsig} \, .
\end{align}
This smoothing further suppresses the feedback of high frequencies between $\atau$ and the body shape.

@@

These fixed-area simulations enable us to precisely characterize the changing geometry as it approaches the terminal state. In particular, we measure the aspect ratio of the body (the width to height) and the opening angle of the front/rear corners. In these tests, we keep the spatial resolution fixed, $N_\iin = 1024$, and show three different cases of the smoothing parameters, $\eps$ and $\sigma$, to assess how the numerical regularization influences shape. In the three cases, we keep $\eps = \sigma$ always and set them to the values 20/1024, 10/1024, and 5/1024. Loosely, these values correspond to smoothing over the nearest 20, 10, and 5 grid points, respectively. Figure~\ref{fig:arangle}(a) shows the aspect ratio as it varies in time. The initially circular geometry has an aspect ratio of unity, and, as erosion carves a more slender form, the aspect ratio increases monotonically. The three simulations are all very close to one another, demonstrating convergence as $\eps, \sigma \to 0$. In the simulation with the least smoothing ($\eps = \sigma = 5/1024$), the aspect ratio tends to a value of 2.7 in long time.

@@

\subsection{Convergence test}
As a final validation, we now verify that our method converges with the expected second-order accuracy in time. For this, we run the same single-body simulation shown in Figs.~\ref{01bodseq} and~\ref{shrink_intface} (except with a coarser $N_\iin=256$) to a stopping time of $t_s = 10^{-2}$. This stopping time is 56\% of the vanishing time, and was selected so that the body will have eroded substantially but is not small enough to compromise accuracy (due to the stress diverging).  We assess error by measuring the $L^2$-difference in the body shape obtained from simulations of successive resolution, with $\Delta t$ halved at each stage. In Table~\ref{convtab}, we report these errors along with the order of accuracy, obtained by Richardson-extrapolation-type calculations. The table confirms that our method converges with expected second-order accuracy. The last column of the table shows the CPU time taken for the simulation to run on a single-processor machine, demonstrating a nearly linear increase in computational time. Though the main parts of the algorithm are easily paralyzed, the single-processor times provide a simple assessment of the overall complexity.

@@

Looking back at the 50-body simulation in Fig.~\ref{fig:50bodies}, many of these same features can be observed. Bodies tend to develop corner-like features connected by flat faces, which causes them to appear polygonal. Nearly straight channels form between the bodies, and a few of these appear to be preferred channels which transmit the most flow and therefore widen the fastest.

@@

To further quantify these observations, Fig.~\ref{fig:mbodyplots} shows several measurements taken from the multibody simulations. In all cases, the black curves correspond to the above 6-body simulation and gray curves to the 50-body simulation. First, Fig.~\ref{fig:mbodyplots}(a) shows the fraction of area occupied by the solid bodies as it varies in time. In the 6-body simulation, the initial configuration comprises about 40\% of the domain area, and this percentage steadily decreases as the bodies erode. The 50-body simulation shows a similar trend.

Next, Figure~\ref{fig:mbodyplots}(b) shows how the flow rate, $\umax$, changes in time as a result of fixing the pressure drop across the cell. Notice that, in this figure, the vertical axis is logarithmic. Both simulations show the flow rate to increase rapidly, even during the early stages of erosion when the area fraction has not changed that significantly. Due to the logarithmic axis, the nearly linear trend seen in the figure corresponds to exponential growth of $\umax$. The change in area fraction alone cannot account for this rapid growth, suggesting the reshaping process to be vital. In particular, the formation of channels observed in Fig.~\ref{fig:06bodies} may account for these higher flow rates.

Naturally, the flow transmitted by the medium is tied to the amount of drag the bodies exert. We therefore show in Fig.~\ref{fig:mbodyplots}(c) the total drag exerted by all bodies as it varies in time. In both cases, the drag remains nearly constant throughout much of the simulation and then drops to zero in the final moments. Although all bodies are diminishing in size, the associated increase in $\umax$ leads to the nearly constant drag that is observed here.
